---
title: "Problem Set 3"
author: "Andrea Moreno"
date: "2026-04-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)
library(knitr)
```

```{r load-data}
compas <- read_csv("data/compas_clean.csv")
```

```{r 1.3a-recidivism-rate}
compas |>
  summarize(overall_recidivism_rate = mean(recidivism, na.rm = TRUE))
```
1.2 
The dataset contains 7,214 observations and 17 variables. The unit of observation is an individual person who was assessed by the COMPAS algorithm.

1.3
a)The overall recidivism rate is approximately 0.45 (or 45.07%).
b)This outcome is binary (it's either a 0 or a 1). This matters for a linear model because the model's answer is a probability and a straight line doesn't know when to stop. It can give you "impossible" answers like 110% or -5% chance of rearrest, which don't make sense in the real world.

```{r 1.4-summary-stats}
# Create a summary table for all variables
compas_summary <- compas |>
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarize(
    Mean = mean(value, na.rm = TRUE),
    SD = sd(value, na.rm = TRUE)
  )


kable(compas_summary, digits = 3, caption = "Summary Statistics for COMPAS Dataset")

```

1.4
Most variation: The variable age has the most variation by far (SD = 11.89). This makes sense because people in the data range from teenagers to seniors. priors_count also has high variation (SD = 4.88) because some people have zero priors while others have many.
Least variation: The variables for native_american (SD = 0.05) and asian (SD = 0.07) have the least variation. This is because almost everyone in this specific dataset is coded as "0" for these categories, so there is very little difference between the observations.

```{r 1.5-recidivism-by-priors}

compas_priors <- compas |>
  mutate(priors_cat = case_when(
    priors_count == 0 ~ "0",
    priors_count >= 1 & priors_count <= 2 ~ "1-2",
    priors_count >= 3 & priors_count <= 5 ~ "3-5",
    priors_count >= 6 & priors_count <= 10 ~ "6-10",
    priors_count >= 11 ~ "11+"
  )) |>
  
  mutate(priors_cat = fct_relevel(priors_cat, "0", "1-2", "3-5", "6-10", "11+"))


compas_priors |>
  group_by(priors_cat) |>
  summarize(rate = mean(recidivism)) |>
  ggplot(aes(x = priors_cat, y = rate, fill = priors_cat)) +
  geom_col() +
  labs(title = "Recidivism Rate by Number of Prior Offenses",
       x = "Number of Priors",
       y = "Recidivism Rate (Proportion)") +
  theme_minimal() +
  guides(fill = "none") # Removes the redundant legend
```
1.5
There is a strong positive relationship between the number of prior offenses and the recidivism rate. As the number of priors increases, the proportion of people who are rearrested within two years goes up steadily—from about 30% for those with zero priors to over 70% for those with 11 or more.

```{r 2.1-split-data}

set.seed(51)

n <- nrow(compas)

train_idx <- sample(1:n, size = floor(0.8 * n))

train <- compas[train_idx, ]
test <- compas[-train_idx, ]

nrow(train)
nrow(test)
```

2.1 
There are 5,771 observations in the training set and 1,443 observations in the test set.
```{r 2.2-ols-model}
ols_model <- lm(recidivism ~ ., data = train)

summary(ols_model)

```
2.2 
a) The predictors that are statistically significant at the 5% level are: age, female, priors_count, felony, has_juv_misd, age_25_45.
b) For every additional prior offense a person has, their predicted probability of recidivism increases by approximately 3 percentage points, holding all other variables constant.

```{r 2.3-rmse-calc}
#in sample
y_hat_train <- predict(ols_model, newdata = train)
rmse_in <- sqrt(mean((y_hat_train - train$recidivism)^2))

#out of sample
y_hat_test <- predict(ols_model, newdata = test)
rmse_out <- sqrt(mean((y_hat_test - test$recidivism)^2))

print(paste("In-sample RMSE:", round(rmse_in, 4)))
print(paste("Out-of-sample RMSE:", round(rmse_out, 4)))
```
2.3
a) In sample: .4606
b) Out of sample: .4622
c)The gap between the in-sample and out-of-sample RMSE is extremely small (0.0016). This tells us that OLS is not overfitting.

```{r 2.4}

set.seed(51)
folds <- sample(rep(1:5, length.out = nrow(train)))
rmse_cv <- numeric(5)

for (k in 1:5) {
  
  train_k <- train[folds != k, ]
  test_k  <- train[folds == k, ]
  
  model_k <- lm(recidivism ~ ., data = train_k)
 
  pred_k  <- predict(model_k, newdata = test_k)
  
  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))
}

mean(rmse_cv)

```
2.4
a) The 5-fold CV estimate of RMSE is 0.4620. This is nearly identical to my single-split out-of-sample RMSE of 0.4622. Because these numbers are so close, we can be very confident that our error estimate is accurate and not just a lucky result from one specific split.
b) Cross-validation is more reliable because it prevents us from getting lucky or unlucky with a single random split. By splitting the data five different times and averaging the results, we ensure that our error estimate isn't being driven by a few weird outliers that happened to land in the test set. It gives us a more stable and honest look at how the model performs on average.
c) Cross-validation is the right tool for choosing lambda because it allows us to test different levels of model complexity. Since lambda controls how much the model penalizes extra variables, CV helps us find the exact spot where the model is complex enough to predict well but simple enough that it doesn't overfit to noise.


