---
title: "Problem Set 3"
author: "Andrea Moreno"
date: "2026-04-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)
library(knitr)
```

```{r load-data}
compas <- read_csv("data/compas_clean.csv")
```

```{r 1.3a-recidivism-rate}
compas |>
  summarize(overall_recidivism_rate = mean(recidivism, na.rm = TRUE))
```
1.2 
The dataset contains 7,214 observations and 17 variables. The unit of observation is an individual person who was assessed by the COMPAS algorithm.

1.3
a)The overall recidivism rate is approximately 0.45 (or 45.07%).
b)This outcome is binary (it's either a 0 or a 1). This matters for a linear model because the model's answer is a probability and a straight line doesn't know when to stop. It can give you "impossible" answers like 110% or -5% chance of rearrest, which don't make sense in the real world.

```{r 1.4-summary-stats}
# Create a summary table for all variables
compas_summary <- compas |>
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarize(
    Mean = mean(value, na.rm = TRUE),
    SD = sd(value, na.rm = TRUE)
  )


kable(compas_summary, digits = 3, caption = "Summary Statistics for COMPAS Dataset")

```

1.4
Most variation: The variable age has the most variation by far (SD = 11.89). This makes sense because people in the data range from teenagers to seniors. priors_count also has high variation (SD = 4.88) because some people have zero priors while others have many.
Least variation: The variables for native_american (SD = 0.05) and asian (SD = 0.07) have the least variation. This is because almost everyone in this specific dataset is coded as "0" for these categories, so there is very little difference between the observations.

```{r 1.5-recidivism-by-priors}

compas_priors <- compas |>
  mutate(priors_cat = case_when(
    priors_count == 0 ~ "0",
    priors_count >= 1 & priors_count <= 2 ~ "1-2",
    priors_count >= 3 & priors_count <= 5 ~ "3-5",
    priors_count >= 6 & priors_count <= 10 ~ "6-10",
    priors_count >= 11 ~ "11+"
  )) |>
  
  mutate(priors_cat = fct_relevel(priors_cat, "0", "1-2", "3-5", "6-10", "11+"))


compas_priors |>
  group_by(priors_cat) |>
  summarize(rate = mean(recidivism)) |>
  ggplot(aes(x = priors_cat, y = rate, fill = priors_cat)) +
  geom_col() +
  labs(title = "Recidivism Rate by Number of Prior Offenses",
       x = "Number of Priors",
       y = "Recidivism Rate (Proportion)") +
  theme_minimal() +
  guides(fill = "none") # Removes the redundant legend
```
1.5
There is a strong positive relationship between the number of prior offenses and the recidivism rate. As the number of priors increases, the proportion of people who are rearrested within two years goes up steadily—from about 30% for those with zero priors to over 70% for those with 11 or more.

```{r 2.1-split-data}

set.seed(51)

n <- nrow(compas)

train_idx <- sample(1:n, size = floor(0.8 * n))

train <- compas[train_idx, ]
test <- compas[-train_idx, ]

nrow(train)
nrow(test)
```

2.1 
There are 5,771 observations in the training set and 1,443 observations in the test set.
```{r 2.2-ols-model}
ols_model <- lm(recidivism ~ ., data = train)

summary(ols_model)

```
2.2 
a) The predictors that are statistically significant at the 5% level are: age, female, priors_count, felony, has_juv_misd, age_25_45.
b) For every additional prior offense a person has, their predicted probability of recidivism increases by approximately 3 percentage points, holding all other variables constant.

```{r 2.3-rmse-calc}
#in sample
y_hat_train <- predict(ols_model, newdata = train)
rmse_in <- sqrt(mean((y_hat_train - train$recidivism)^2))

#out of sample
y_hat_test <- predict(ols_model, newdata = test)
rmse_out <- sqrt(mean((y_hat_test - test$recidivism)^2))

print(paste("In-sample RMSE:", round(rmse_in, 4)))
print(paste("Out-of-sample RMSE:", round(rmse_out, 4)))
```
2.3
a) In sample: .4606
b) Out of sample: .4622
c)The gap between the in-sample and out-of-sample RMSE is extremely small (0.0016). This tells us that OLS is not overfitting.

```{r 2.4}

set.seed(51)
folds <- sample(rep(1:5, length.out = nrow(train)))
rmse_cv <- numeric(5)

for (k in 1:5) {
  
  train_k <- train[folds != k, ]
  test_k  <- train[folds == k, ]
  
  model_k <- lm(recidivism ~ ., data = train_k)
 
  pred_k  <- predict(model_k, newdata = test_k)
  
  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))
}

mean(rmse_cv)

```
2.4
a) The 5-fold CV estimate of RMSE is 0.4620. This is nearly identical to my single-split out-of-sample RMSE of 0.4622. Because these numbers are so close, we can be very confident that our error estimate is accurate and not just a lucky result from one specific split.
b) Cross-validation is more reliable because it prevents us from getting lucky or unlucky with a single random split. By splitting the data five different times and averaging the results, we ensure that our error estimate isn't being driven by a few weird outliers that happened to land in the test set. It gives us a more stable and honest look at how the model performs on average.
c) Cross-validation is the right tool for choosing lambda because it allows us to test different levels of model complexity. Since lambda controls how much the model penalizes extra variables, CV helps us find the exact spot where the model is complex enough to predict well but simple enough that it doesn't overfit to noise.

```{r 3.1}

f_full <- recidivism ~ (.)^2 + 
  I(age^2) + I(priors_count^2) + 
  I(juv_fel_count^2) + I(juv_misd_count^2) + 
  I(juv_other_count^2)

X_train <- model.matrix(f_full, data = train)[, -1]
X_test  <- model.matrix(f_full, data = test)[, -1]

y_train <- train$recidivism
y_test  <- test$recidivism

ncol(X_train)

nrow(train) / ncol(X_train)
```
3.1
a) The expanded model has 141 predictors.
b) The ratio of training observations to predictors is now approximately 41-to-1 (5,771 / 141). In our first model, we had about 360 rows of data for every 1 variable we tried to estimate. Now, we only have about 41 rows for every 1 variable.

```{r 3.2}

ols_full <- lm(f_full, data = train)

y_hat_train_full <- predict(ols_full, newdata = train)
rmse_in_full <- sqrt(mean((y_hat_train_full - train$recidivism)^2))
y_hat_test_full <- predict(ols_full, newdata = test)
rmse_out_full <- sqrt(mean((y_hat_test_full - test$recidivism)^2))

print(paste("Full Model In-sample RMSE:", round(rmse_in_full, 4)))
print(paste("Full Model Out-of-sample RMSE:", round(rmse_out_full, 4)))
```
3.2 
a) in sample RMSE: 0.4485
b) out of sample: 0.489
c) The gap has grown significantly compared to the simple model where the gap was only 0.0016.The In sample RMSE dropped (from 0.4606 to 0.4485) because the model used those 141 predictors to memorize the specific noise and random quirks of the training data. The out of sample RMSE increased (from 0.4622 to 0.4890) because the noise it memorized in the training set doesn't exist in the test set, the model's predictions actually got worse for new people.

3.3 
This phenomenon is called overfitting.By adding so many interaction terms, the model was able to memorize the specific noise and random coincidences in the training data, which made the in sample fit appear better. However, because those random patterns do not exist in the test data, the model’s out of sample predictions got worse as it tried to apply fake rules to new people.

```{r 4.1}
library(glmnet)
set.seed(51)
cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0)
best_lambda_ridge <- cv_ridge$lambda.min
print(paste("Optimal Ridge Lambda:", round(best_lambda_ridge, 4)))
plot(cv_ridge)
```
4.2
a) Optimal lambda: .0138
b) As we move from left to right, increasing the penalty, the mean squared error drops. This shows that the original OLS model was overfitting. By shrinking the coefficients, we made the model much more accurate. Also, in Ridge regression, the number of predictors stays constant. Even though we are shrinking them, Ridge never sets a coefficient exactly to zero. This is why 128 is across the top, all variables are still in the model.

```{r 4.2}
y_hat_ridge <- predict(cv_ridge, newx = X_test, s = "lambda.min")

rmse_ridge <- sqrt(mean((y_hat_ridge - y_test)^2))

print(paste("Simple OLS RMSE (16 vars):", 0.4622))
print(paste("Kitchen-Sink OLS RMSE (141 vars):", 0.4890))
print(paste("Ridge RMSE (141 vars + Penalty):", round(rmse_ridge, 4)))
```
4.2 
The Ridge RMSE (0.4700) is an improvement over the Kitchen-Sink OLS (0.4890), but it still performs worse than the original Simple OLS (0.4622). Ridge added a penalty that tamed the 141 variables, which prevented the overfitting from earlier. This is why the error dropped from 0.489 to 0.470. Even with the Ridge penalty, the 141 interaction and squared terms might be introducing more noise than signal. Because ridge keeps all 141 variables in the model it never sets them to zero, it is still carrying around a lot of dead weight that hurts its ability to predict as accurately as the simple model.

```{r 4.3}
ridge_coefs <- coef(cv_ridge, s = "lambda.min")
sum(ridge_coefs == 0)
length(ridge_coefs)
```
4.3
No ridge regression did not perform any active feature selection. While the output shows 13 coefficients as zero, these were dropped due to linear redundancy in the Kitchen Sink data. Ridge uses an L2 penalty, which squares the coefficients in the optimization function. This approach shrinks coefficients toward zero to reduce variance, but it never forces them to be exactly zero. Ridge performs shrinkage but not feature selection, keeping all 129 valid predictors in the model.

```{r 5.1}
set.seed(51)
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1)
best_lambda_lasso <- cv_lasso$lambda.min
print(paste("Optimal LASSO Lambda:", round(best_lambda_lasso, 4)))

plot(cv_lasso)
```

5.1
a)0.0077
b)The plot shows how the mean squared error changes as the penalty increases. The numbers across the top of this graph decrease from right to left, showing that LASSO is actively selecting variables by setting their coefficients to exactly zero.

```{r 5.2}
lasso_coefs <- as.matrix(coef(cv_lasso, s = "lambda.min"))
survived <- sum(lasso_coefs != 0) - 1  # -1 to exclude the Intercept
eliminated <- sum(lasso_coefs == 0)

print(paste("Variables that survived:", survived))
print(paste("Variables eliminated:", eliminated))
head(lasso_coefs[lasso_coefs != 0 & grepl(":", rownames(lasso_coefs)), , drop=FALSE], 10)
```
5.2
a) Out of the original ~141 variables, 27 survived with non-zero coefficients. LASSO eliminated 114 variables.
b) A few surviving interaction terms include female:age_under_25 and black:age_under_25.Yes, these make sense. They suggest that the baselinen effect of race or gender on recidivism is not the same for everyone, it might be different specifically for younger individuals. By keeping these, the model is acknowledging that the relationship between these demographics and the outcome is more complex than what our simple OLS model assumed.
c) What we see here illustrates his point the computer, not the researcher, made the choice. Instead of me guessing whether female:priors_count was important, the LASSO algorithm used a strict mathematical penalty to decide. This removes the researcher's subjective discretion and replaces it with a transparent, data driven rule for which variables belong in the final report.

```{r 5.3}
y_hat_lasso <- predict(cv_lasso, newx = X_test, s = "lambda.min")
rmse_lasso <- sqrt(mean((y_hat_lasso - y_test)^2))
results <- data.frame(
  Model = c("Simple OLS (16 vars)", "Kitchen-Sink OLS (141 vars)", "Ridge (Shrinkage)", "LASSO (Selection)"),
  RMSE = c(0.4622, 0.4890, 0.4700, round(rmse_lasso, 4))
)

print(results)
```
5.3
Compared to - 
OLS (0.4622): LASSO performed the best. It proved that there was useful information in those extra 141 variables (like the interactions between age and race), but you needed to filter out the garbage to see the benefit.

Kitchen sink OLS (0.4890): LASSO massively improved the model. It reduced the error by over 6%, showing that more data is only better if you have a way to handle the overfitting.

Ridge (0.4700): LASSO performed better than Ridge. In this specific recidivism dataset, it was more effective to completely zero out the weak predictors than to just shrink all of them. This suggests that a few key interactions matter a lot, while most of the others are truly useless.

```{r 6.1}
set.seed(51)
cv_enet <- cv.glmnet(X_train, y_train, alpha = 0.5)

y_hat_enet <- predict(cv_enet, newx = X_test, s = "lambda.min")
rmse_enet <- sqrt(mean((y_hat_enet - y_test)^2))
enet_coefs <- as.matrix(coef(cv_enet, s = "lambda.min"))
retained_enet <- sum(enet_coefs != 0) - 1 # Exclude intercept

print(paste("Elastic Net RMSE:", round(rmse_enet, 4)))
print(paste("Variables Retained:", retained_enet))
```
6.1 
Out of sample RMSE:  0.4583
Variables Retained: 43

```{r 6.2}
library(knitr)
comparison_df <- data.frame(
  Model = c("OLS (main effects)", 
            "OLS (kitchen sink)", 
            "Ridge", 
            "LASSO", 
            "Elastic Net"),
  Predictors = c(16, 141, 128, 27, 43),
  `Out-of-Sample RMSE` = c(0.4622, 0.4890, 0.4700, 0.4593, 0.4583)
)


kable(comparison_df, 
      digits = 4,
      align = "lcc")
```

6.2
a) The Elastic Net model has the lowest out-of-sample RMSE (0.4583).
b) The Kitchen-Sink OLS has the best in-sample fit. It is not the same model, its high performance in-sample was actually a sign of overfitting, which is why it performed the worst out-of-sample.
c) This table shows that simply adding more variables (moving from 16 to 141) increases variance so much that the model's predictive power collapses. Regularization techniques like LASSO and elastic net intentionally introduce a little bit of bias by shrinking coefficients, but in return, they significantly reduce variance. This tradeoff allows them to achieve a much lower RMSE than the standard OLS models.

6.3
a) LASSO helped narrow the disagreement by removing subjective discretion from the model-building process. In the Halliburton case, experts could cherry pick specific peer firms to include in their OLS models to tip the results in their favor. By using LASSO, the algorithm determines which variables or peer firms are statistically relevant based on a strict, mathematical penalty. This creates a transparent rule that prevents both sides from manually selecting only the data that supports their desired conclusion.
b) Based on this problem set, the manual selection of interaction terms and control variables gives the analyst the most room for manipulation. In the Kitchen Sink model, we saw how easily 141 variables can be created. In a traditional OLS approach, an analyst could trial-and-error hundreds of combinations of these interactions and only report the one that produces a significant p-value or a specific coefficient. Unlike LASSO, which forces a penalty on every extra variable, traditional OLS allows a researcher to hide the noise and present a hand-picked model as if it were the only one they tested.

6.4
a)No, it would not. The LASSO coefficient is biased by design. Because LASSO uses a penalty to shrink coefficients toward zero to improve prediction, the resulting number is not an unbiased estimate of the true effect of prior offenses. Furthermore, LASSO might mute other variables that are technically related to priors_count (confounders), which would normally be required to isolate a causal effect. LASSO cares about accuracy, not honesty regarding the underlying mechanism.
b) The goal in using regression for prediction is to minimize the distance between our guess and the actual outcome. Any variable can help predict recidivism, a predictive model would use it. It is focused on correlation and out-of-sample performance. The goal in using it for casual inference is to isolate the effect of a specific treatment while holding everything else constant. This requires a strict theoretical model to account for all confounding factors. A causal model might have a worse RMSE than a LASSO model, but it provides a more reliable answer to If we change X, will Y change?


```{r 6.5}
is_white <- test$black == 0 & test$hispanic == 0 & 
            test$asian == 0 & test$other_race == 0 & 
            test$native_american == 0

is_black <- test$black == 1
y_hat_simple <- predict(ols_model, newdata = test)
mean_pred_white <- mean(y_hat_simple[is_white], na.rm = TRUE)
mean_pred_black <- mean(y_hat_simple[is_black], na.rm = TRUE)
mean_actual_white <- mean(test$recidivism[is_white], na.rm = TRUE)
mean_actual_black <- mean(test$recidivism[is_black], na.rm = TRUE)

results_fairness <- data.frame(
  Group = c("White", "Black"),
  `Predicted Rate` = c(mean_pred_white, mean_pred_black),
  `Actual Rate` = c(mean_actual_white, mean_actual_black)
)
print(results_fairness)
```
6.5
a) The mean predicted recidivism probability for Black defendants is 0.512, while for white defendants it is 0.379. There is a difference of roughly 13 percentage points in how the model assesses risk based on race.
b) The actual recidivism rate for Black defendants is 0.521 and for white defendants is 0.402. The model’s predictions are remarkably close to the actual rates in the test set, meaning the model is accurately capturing the disparities present in the data
c) A tradeoff in using algorithms like COMPAS is between predictive accuracy and systemic equity. While the model is accurate in that its predictions match the actual rates, it relies on variables like priors_count which often reflect historical over-policing of black communities. By using these statistically significant predictors to lower RMSE, the algorithm risks automating and validating racial disparities in the justice system rather than offering a neutral assessment of individual risk.

