---
title: "Problem Set 3"
author: "Andrea Moreno"
date: "2026-04-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)
library(knitr)
```

```{r load-data}
compas <- read_csv("data/compas_clean.csv")
```

```{r 1.3a-recidivism-rate}
compas |>
  summarize(overall_recidivism_rate = mean(recidivism, na.rm = TRUE))
```
1.2 
The dataset contains 7,214 observations and 17 variables. The unit of observation is an individual person who was assessed by the COMPAS algorithm.

1.3
a)The overall recidivism rate is approximately 0.45 (or 45.07%).
b)This outcome is binary (it's either a 0 or a 1). This matters for a linear model because the model's answer is a probability and a straight line doesn't know when to stop. It can give you "impossible" answers like 110% or -5% chance of rearrest, which don't make sense in the real world.

```{r 1.4-summary-stats}
# Create a summary table for all variables
compas_summary <- compas |>
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarize(
    Mean = mean(value, na.rm = TRUE),
    SD = sd(value, na.rm = TRUE)
  )


kable(compas_summary, digits = 3, caption = "Summary Statistics for COMPAS Dataset")

```

1.4
Most variation: The variable age has the most variation by far (SD = 11.89). This makes sense because people in the data range from teenagers to seniors. priors_count also has high variation (SD = 4.88) because some people have zero priors while others have many.
Least variation: The variables for native_american (SD = 0.05) and asian (SD = 0.07) have the least variation. This is because almost everyone in this specific dataset is coded as "0" for these categories, so there is very little difference between the observations.

```{r 1.5-recidivism-by-priors}

compas_priors <- compas |>
  mutate(priors_cat = case_when(
    priors_count == 0 ~ "0",
    priors_count >= 1 & priors_count <= 2 ~ "1-2",
    priors_count >= 3 & priors_count <= 5 ~ "3-5",
    priors_count >= 6 & priors_count <= 10 ~ "6-10",
    priors_count >= 11 ~ "11+"
  )) |>
  
  mutate(priors_cat = fct_relevel(priors_cat, "0", "1-2", "3-5", "6-10", "11+"))


compas_priors |>
  group_by(priors_cat) |>
  summarize(rate = mean(recidivism)) |>
  ggplot(aes(x = priors_cat, y = rate, fill = priors_cat)) +
  geom_col() +
  labs(title = "Recidivism Rate by Number of Prior Offenses",
       x = "Number of Priors",
       y = "Recidivism Rate (Proportion)") +
  theme_minimal() +
  guides(fill = "none") # Removes the redundant legend
```
1.5
There is a strong positive relationship between the number of prior offenses and the recidivism rate. As the number of priors increases, the proportion of people who are rearrested within two years goes up steadily—from about 30% for those with zero priors to over 70% for those with 11 or more.


