---
title: "Problem Set 3"
author: "Andrea Moreno"
date: "2026-04-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)
library(knitr)
```

```{r load-data}
compas <- read_csv("data/compas_clean.csv")
```

```{r 1.3a-recidivism-rate}
compas |>
  summarize(overall_recidivism_rate = mean(recidivism, na.rm = TRUE))
```
1.2 
The dataset contains 7,214 observations and 17 variables. The unit of observation is an individual person who was assessed by the COMPAS algorithm.

1.3
a)The overall recidivism rate is approximately 0.45 (or 45.07%).
b)This outcome is binary (it's either a 0 or a 1). This matters for a linear model because the model's answer is a probability and a straight line doesn't know when to stop. It can give you "impossible" answers like 110% or -5% chance of rearrest, which don't make sense in the real world.

```{r 1.4-summary-stats}
# Create a summary table for all variables
compas_summary <- compas |>
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarize(
    Mean = mean(value, na.rm = TRUE),
    SD = sd(value, na.rm = TRUE)
  )


kable(compas_summary, digits = 3, caption = "Summary Statistics for COMPAS Dataset")

```

1.4
Most variation: The variable age has the most variation by far (SD = 11.89). This makes sense because people in the data range from teenagers to seniors. priors_count also has high variation (SD = 4.88) because some people have zero priors while others have many.
Least variation: The variables for native_american (SD = 0.05) and asian (SD = 0.07) have the least variation. This is because almost everyone in this specific dataset is coded as "0" for these categories, so there is very little difference between the observations.

```{r 1.5-recidivism-by-priors}

compas_priors <- compas |>
  mutate(priors_cat = case_when(
    priors_count == 0 ~ "0",
    priors_count >= 1 & priors_count <= 2 ~ "1-2",
    priors_count >= 3 & priors_count <= 5 ~ "3-5",
    priors_count >= 6 & priors_count <= 10 ~ "6-10",
    priors_count >= 11 ~ "11+"
  )) |>
  
  mutate(priors_cat = fct_relevel(priors_cat, "0", "1-2", "3-5", "6-10", "11+"))


compas_priors |>
  group_by(priors_cat) |>
  summarize(rate = mean(recidivism)) |>
  ggplot(aes(x = priors_cat, y = rate, fill = priors_cat)) +
  geom_col() +
  labs(title = "Recidivism Rate by Number of Prior Offenses",
       x = "Number of Priors",
       y = "Recidivism Rate (Proportion)") +
  theme_minimal() +
  guides(fill = "none") # Removes the redundant legend
```
1.5
There is a strong positive relationship between the number of prior offenses and the recidivism rate. As the number of priors increases, the proportion of people who are rearrested within two years goes up steadily—from about 30% for those with zero priors to over 70% for those with 11 or more.

```{r 2.1-split-data}

set.seed(51)

n <- nrow(compas)

train_idx <- sample(1:n, size = floor(0.8 * n))

train <- compas[train_idx, ]
test <- compas[-train_idx, ]

nrow(train)
nrow(test)
```

2.1 
There are 5,771 observations in the training set and 1,443 observations in the test set.
```{r 2.2-ols-model}
ols_model <- lm(recidivism ~ ., data = train)

summary(ols_model)

```
2.2 
a) The predictors that are statistically significant at the 5% level are: age, female, priors_count, felony, has_juv_misd, age_25_45.
b) For every additional prior offense a person has, their predicted probability of recidivism increases by approximately 3 percentage points, holding all other variables constant.

```{r 2.3-rmse-calc}
#in sample
y_hat_train <- predict(ols_model, newdata = train)
rmse_in <- sqrt(mean((y_hat_train - train$recidivism)^2))

#out of sample
y_hat_test <- predict(ols_model, newdata = test)
rmse_out <- sqrt(mean((y_hat_test - test$recidivism)^2))

print(paste("In-sample RMSE:", round(rmse_in, 4)))
print(paste("Out-of-sample RMSE:", round(rmse_out, 4)))
```
2.3
a) In sample: .4606
b) Out of sample: .4622
c)The gap between the in-sample and out-of-sample RMSE is extremely small (0.0016). This tells us that OLS is not overfitting.

```{r 2.4}

set.seed(51)
folds <- sample(rep(1:5, length.out = nrow(train)))
rmse_cv <- numeric(5)

for (k in 1:5) {
  
  train_k <- train[folds != k, ]
  test_k  <- train[folds == k, ]
  
  model_k <- lm(recidivism ~ ., data = train_k)
 
  pred_k  <- predict(model_k, newdata = test_k)
  
  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))
}

mean(rmse_cv)

```
2.4
a) The 5-fold CV estimate of RMSE is 0.4620. This is nearly identical to my single-split out-of-sample RMSE of 0.4622. Because these numbers are so close, we can be very confident that our error estimate is accurate and not just a lucky result from one specific split.
b) Cross-validation is more reliable because it prevents us from getting lucky or unlucky with a single random split. By splitting the data five different times and averaging the results, we ensure that our error estimate isn't being driven by a few weird outliers that happened to land in the test set. It gives us a more stable and honest look at how the model performs on average.
c) Cross-validation is the right tool for choosing lambda because it allows us to test different levels of model complexity. Since lambda controls how much the model penalizes extra variables, CV helps us find the exact spot where the model is complex enough to predict well but simple enough that it doesn't overfit to noise.

```{r 3.1}

f_full <- recidivism ~ (.)^2 + 
  I(age^2) + I(priors_count^2) + 
  I(juv_fel_count^2) + I(juv_misd_count^2) + 
  I(juv_other_count^2)

X_train <- model.matrix(f_full, data = train)[, -1]
X_test  <- model.matrix(f_full, data = test)[, -1]

y_train <- train$recidivism
y_test  <- test$recidivism

ncol(X_train)

nrow(train) / ncol(X_train)
```
3.1
a) The expanded model has 141 predictors.
b) The ratio of training observations to predictors is now approximately 41-to-1 (5,771 / 141). In our first model, we had about 360 rows of data for every 1 variable we tried to estimate. Now, we only have about 41 rows for every 1 variable.

```{r 3.2}

ols_full <- lm(f_full, data = train)

y_hat_train_full <- predict(ols_full, newdata = train)
rmse_in_full <- sqrt(mean((y_hat_train_full - train$recidivism)^2))
y_hat_test_full <- predict(ols_full, newdata = test)
rmse_out_full <- sqrt(mean((y_hat_test_full - test$recidivism)^2))

print(paste("Full Model In-sample RMSE:", round(rmse_in_full, 4)))
print(paste("Full Model Out-of-sample RMSE:", round(rmse_out_full, 4)))
```
3.2 
a) in sample RMSE: 0.4485
b) out of sample: 0.489
c) The gap has grown significantly compared to the simple model where the gap was only 0.0016.The In sample RMSE dropped (from 0.4606 to 0.4485) because the model used those 141 predictors to memorize the specific noise and random quirks of the training data. The out of sample RMSE increased (from 0.4622 to 0.4890) because the noise it memorized in the training set doesn't exist in the test set, the model's predictions actually got worse for new people.

3.3 
This phenomenon is called overfitting.By adding so many interaction terms, the model was able to memorize the specific noise and random coincidences in the training data, which made the in sample fit appear better. However, because those random patterns do not exist in the test data, the model’s out of sample predictions got worse as it tried to apply fake rules to new people.

```{r 4.1}
library(glmnet)
set.seed(51)
cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0)
best_lambda_ridge <- cv_ridge$lambda.min
print(paste("Optimal Ridge Lambda:", round(best_lambda_ridge, 4)))
plot(cv_ridge)
```
4.2
a) Optimal lambda: .0138
b) As you move from left to right, increasing the penalty, the mean squared error drops. This shows that the original OLS model was overfitting. By shrinking the coefficients, we made the model much more accurate. Also, in Ridge regression, the number of predictors stays constant. Even though we are shrinking them, Ridge never sets a coefficient exactly to zero. This is why 128 is across the top, all variables are still in the model.

```{r 4.2}
y_hat_ridge <- predict(cv_ridge, newx = X_test, s = "lambda.min")

rmse_ridge <- sqrt(mean((y_hat_ridge - y_test)^2))

print(paste("Simple OLS RMSE (16 vars):", 0.4622))
print(paste("Kitchen-Sink OLS RMSE (141 vars):", 0.4890))
print(paste("Ridge RMSE (141 vars + Penalty):", round(rmse_ridge, 4)))
```
4.2 
The Ridge RMSE (0.4700) is an improvement over the Kitchen-Sink OLS (0.4890), but it still performs worse than the original Simple OLS (0.4622). Ridge added a penalty that tamed the 141 variables, which prevented the overfitting from earlier. This is why the error dropped from 0.489 to 0.470. Even with the Ridge penalty, the 141 interaction and squared terms might be introducing more noise than signal. Because ridge keeps all 141 variables in the model it never sets them to zero, it is still carrying around a lot of dead weight that hurts its ability to predict as accurately as the simple model.

```{r 4.3}
ridge_coefs <- coef(cv_ridge, s = "lambda.min")
sum(ridge_coefs == 0)
length(ridge_coefs)
```
4.3
No ridge regression did not perform any active feature selection. While the output shows 13 coefficients as zero, these were dropped due to linear redundancy in the Kitchen Sink data. Ridge uses an L2 penalty, which squares the coefficients in the optimization function. This approach shrinks coefficients toward zero to reduce variance, but it never forces them to be exactly zero. Ridge performs shrinkage but not feature selection, keeping all 129 valid predictors in the model.

